---
title: "Generate Inapplicable data"
author: "[Thomas Guillerme](mailto:guillert@tcd.ie)"
date: "`r Sys.Date()`"
output: pdf_document
---

This is a quick example for simulating inapplicable data in R.

Before starting
===============

You'll need `ape`, `phyclust`, `devtools` and `dispRity` packages.
The latter needs to be installed *via* `devtools` as follows:

```{r, eval = FALSE}
if(!require(devtools)) install.packages("devtools")
if(!require(ape)) install.packages("ape")
if(!require(phyclust)) install.packages("phyclust")
install_github("TGuillerme/dispRity", ref = "release")
```

Once installed, everything should be loaded:s

```{r}
library(devtools)
library(ape)
library(phyclust)
library(dispRity)
set.seed(0)
```

Generating the inapplicable characters
===============

Getting the trees
---

You'll need a tree first.
Either an input tree:

```{r, eval = FALSE}
tree <- read.tree(text = "(a,(b,c));")
```
> Note that you can also get a newick tree using `read.tree(file = "myTreeFile.tre")` or a nexus one using `read.nexus(file = "myNexusFile.nex")`.

Or generate a random one with `n` taxa (let's do 20).

```{r}
## Number of taxa
n <- 20
## Tree (birth-death)
tree <- rphylo(n, 0.2, 0.05)
```
Generating the characters
---

The characters can be simulated using `sim.morpho` from the `dispRity` package following couple of parameters (rates, states, etc...).
The idea is to first generate a binary character that will be our ''presence-absence'' one and a second one that will be the states of the present feature:

```{r}
## The overall character rate
overall_rate <- c(rgamma, rate = 5, shape = 10)

## The first character ("presence-absence")
presence_absence <- sim.morpho(tree, characters = 1, rates = overall_rate,
    invariant = FALSE)

## The first character ("presence_states-absence")
presence_states <- sim.morpho(tree, characters = 1, states = c(0.75, 0.15, 0.1),
    rates = overall_rate, invariant = FALSE)

## Combine the two characters
matrix <- matrix(c(presence_absence, presence_states), ncol = 2, byrow = FALSE)
rownames(matrix) <- rownames(presence_states)

```

We can then apply the inapplicable data using the first character (presence/absences):

```{r}
matrix <- apply.inapplicable(matrix, inapplicables = "character")
```

By default, the first character is used for generating the inapplicable data on the second character. Any `0` will be translated into an inapplicable character on the next character.

Visualising the character
---


```{r, fig.align='center',fig.width=8, fig.height=8}
plot(tree, show.tip.label = TRUE, label.offset = 0, cex = 0.7)
tiplabels(pch = 21, bg = c("white", "black")[as.factor(matrix[,1])], cex = 1.1, adj = 1.2)
tiplabels(text = as.character(matrix[,2]), cex = 0.7, adj = -5.4, frame = "none")
```

<!-- matrix[,2] <- gsub("-", "–", matrix[,2]) -->